A Wait-free Queue with Poly-logarithmic Step Complexity

### 2022-11-14 thesis/thesis.tex
apply more comments 2

\vspace{3em}
\textcopyright Hossein Naderibeni, 2022

@@

\section*{Acknowledgements}
\hspace{\parindent}First, I want to thank my supervisor Eric Ruppert. Without his guidance and support, this thesis probably would not have been completed.

I would also like to thank my co-supervisor Franck van Breugel, not only for his careful reading of my thesis and his thoughtful comments bu also for his academic advice and motivations.

@@

Last but not the least, I would like to thank my parents, for providing me with unfailing support and continuous encouragement throughout my years of study. I wanted to thank my brother for being there for me whenever I needed any help.

@@

I dedicate this paper to my people, who are fighting for their freedom.

@@

\pagenumbering{arabic}
\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. It is not hard to see why multiple processes cannot share a sequential data structure. For example, consider two processes trying to append to a sequential linked list simultaneously. Processes $p,q$ read the same tail node, $p$ changes the next pointer of the tail node to its new node, and then $q$ does the same. In this run, $p$'s update is overwritten. One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked. For these reasons, we are interested in the design of an efficient shared queue without using locks.

@@

Nikolaev and Ravindran~\cite{DBLP:journals/corr/abs-2201-02179} present a wait-free queue that uses the fast-path slow-path methodology introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}. Their work is based on a circular queue using bounded memory. When a process wishes to do an enqueue or a dequeue, it starts two paths. The fast path  ensures good performance while the slow path ensures termination. They show that these two paths do not affect each other and the queue remains consistent. If a process makes no progress, other processes help its slow path to finish. The helping phase suffers from the CAS Retry Problem, because processes compete in a \nf{CAS} loop to decide which succeeds to help. Because of this, the worst-case complexity cannot be better than $\Omega(p)$.

@@

\pagebreak
\newpage

\subsection{Pseudocode} \label{algQ}
We present our algorithm in pseudocode. Page~23 contains the description of the fields in the tree nodes and the blocks. Value of any uninitialized field is \nf{null}. Page~24 contains major routines and the rest of this section consists of the auxiliary routines
The abbreviations bellow are used in the pseudocode and the proof of correctness.
\begin{itemize}
 \item \tt{blocks[$b$].sum\sub{x}=blocks[$b$].sum\sub{x-left}+blocks[$b$].sum\sub{x-right}} (for internal blocks where $b\geq 0$ and \tt{x} $\in$ \{enq, deq\})
%  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
  \item \tt{blocks[$b$].num\sub{x}=blocks[$b$].sum\sub{x}-blocks[$b-1$].sum\sub{x}}  (for all blocks where $b>0$ and \tt{x} $\in$ \tt{\{enq, deq, enq-left, enq-right, deq-left, deq-right\}})
\end{itemize}

@@

Now we define the ordering of operations stored in each node. In the non-root nodes, we only need to order operations of a type among themselves that is, we order the \nf{Enqueue}s in the node and order the \nf{Dequeue}s in the node separately. Processes are numbered from 1 to $p$, and leaves of the tree are assigned from left to right. We will show in Lemma \ref{blockSize} that there is at most one operation from each process in a given block.
\begin{definition} [Ordering of operations inside the nodes] \label{ordering}
\end{definition}
\begin{itemize}
  \item $E(n,b)$ is the sequence of enqueue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows. $E(leaf,b)$ is the single enqueue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents a dequeue operation. If $n$ is an internal node, then
\begin{align*} 
E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
\end{align*}
  \item $E_i(n,b)$ is the $i$th enqueue in $E(n,b)$.
\item The order of the enqueue operations in the node $n$ is $E(n)=E(n,1)\cdot E(n,2)\cdot E(n,3)\cdots$
\item $E_i(n)$ is the $i$th enqueue in $E(n)$.
  \item $D(n,b)$ is the sequence of dequeue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows. $D(leaf,b)$ is the single dequeue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents an enqueue operation. If $n$ is an internal node, then
\begin{align*} 
D(n,b) =&  D(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots D(n\nf{.left}, n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
&D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots D(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
\end{align*}
    \item $D_i(n,b)$ is the $i$th enqueue in $D(n,b)$.
\item The order of the dequeue operations in the node $n$ is $D(n)=D(n,1)\cdot D(n,2)\cdot D(n,3)...$
\item $D_i(n)$ is the $i$th dequeue in $D(n)$.
\end{itemize}

@@

\begin{lemma} \label{doubleRefresh}
  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh} on internal node $n$ by process $p$. If neither $R_1$ nor $R_2$ is a successful \nf{Refresh}, then we have $EST_{n\nf{.left}}^{t^{R{_1}}} \; \cup \; EST_{n\nf{.right}}^{t^{R{_1}}} \subseteq EST_{n}^{^{R{_2}}t}$.
\end{lemma}
\begin{proof}
Let $R_1$ read $i$ from $n\nf{.head}$ at Line \ref{readHead}. By Lemma \ref{lem::headInc}, $R_1$ and $R_2$ cannot both read the same value $i$. By Observation \ref{nonDecreasingHead}, $R_2$ reads a larger value of $n$\nf{.head} than $R_1$.

@@

\begin{lemma}\label{blockSize}
Each block contains at most one operation of each process and, therefore, at most $p$ operations in total.
\end{lemma}
\begin{proof}
To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $p$ in block $B$ in node $n$. Without loss of generality $op_1$ is invoked earlier than $op_2$. Process $p$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins. By Corollary \ref{lem::appendExactlyOnce}, before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $p$'s leaf to the root. Since $b$ contains $op_2$, it must be created after $op_2$ is invoked. The fact that $op_2$\nf{.Append} is invoked after $op_1$\nf{.Append} terminated means that there is some block $B^\prime$ in $n$ before $B$ that contains $op_1$. The existence of $op_1$ in $B$ and $B^\prime$ contradicts Lemma~\ref{lem::noDuplicates}.
\end{proof}

\begin{lemma}\label{subBlocksBound}
Each block has at most $p$ direct subblocks.
\end{lemma}
\begin{proof}
From the definition of operations in a block in Definition \ref{def::ops} we know the operations in a block are the union of the operations in the direct subblocks of the block. The claim follows from the equivalence of the number of operations in a block and the number of operations in the subblocks of the block and the observation that each block appended to an internal node contains at least one operation due to the test on Line~\ref{addOP}. Also, blocks in the leaves contain only one \nf{Enqueue} or \nf{Dequeue} operation.
\end{proof}

@@

\begin{lemma}[\nf{DoublingSearch} correctness]\label{dsearch}
If $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, then \nf{Doubling-}\\\nf{Search($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
\end{lemma}
\begin{proof}
 From Lines \ref{setSumEnqLeft} and \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}}, and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order. Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing. By Corollary~\ref{lem::sumX} we know that the \nf{sum\sub{enq}} field in a block is the sum of the number of \nf{Enqueue} operations in that block and the prefix blocks of that block in the node. Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b-1$].sum\sub{enq}}<e\leq\nf{root.blocks[$b$].sum\sub{enq}}$. Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.  Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles the search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$. Then, in Line \ref{dsearchBinarySearch}, the binary search finds the $b$ such that \nf{root.blocks[$b-1$].sum\sub{enq}$<e\leq$root.blocks[$b$].sum\sub{enq}}. By Corollary \ref{lem::sumX}, \nf{root.blocks[$b$]} is the block that contains $E_e(root)$. Finally, $i$ is computed using the definition of \nf{sum\sub{enq}} and Corollary \ref{lem::sumX}.
\end{proof}

@@

  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a right child]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a right child. The index of the superblock is shown with $sb$. }\label{fig::blockSumDeqRight}
\end{figure}
% \%TODO :Change sb to superblockIndex
% \%TODO :Change subblock to subblockIndex in GetEnqueue proof

@@

In the next lemma, we show that the \nf{size} field in each \nf{root block} is computed correctly.
\begin{lemma}\label{sizeCorrectness}
  \nf{root.blocks[$b$].\size}is the \size of the queue after the operations in \nf{root.blocks[$0\cdots b$]} are applied in the order of~$L$.  
\end{lemma}
\begin{proof}
We prove the claim by induction on $b$. The base case when ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{]}$ contains an empty block with \nf{size} field equal to $0$. We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$. By Definition \ref{ordering} \nf{Enqueue} operations come before \nf{Dequeue} operations in a block in $L$. By Lemma~\ref{lem::numX} \nf{num\sub{enq}} and \nf{num\sub{deq}} fields in a block show the number of \nf{Enqueue} and \nf{Dequeue} operations in it. If there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}. Otherwise, the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- \nf{root.blocks[}b\nf{].num\sub{deq}}$. In both cases, this is the same as the assignment on Line \ref{computeLength}.
\end{proof}

@@

\begin{lemma}\label{nullReturn}
$Resp(D_i(\nf{root},b))$ is \nf{null} iff \nf{root.blocks[$b-1$].\size $+$ root.blocks[$b$].num\sub{enq}$- i$ $<0$}.
\end{lemma}
\begin{proof}
Immediately from Corollary \ref{numNullDeqBlock} and Lemma \ref{lem::numX}.
\end{proof}

@@

A \nf{Dequeue} creates a block whose \nf{element} is \nf{null}, appends it to the tree, computes its rank among non-null dequeues, finds the corresponding enqueue and returns the response. The first two parts are similar to an \nf{Enqueue} operation and take $O(\log p)$ steps. To compute the rank of a \nf{Dequeue} in $D(n)$, the \nf{Dequeue} calls \nf{IndexDequeue()}. \nf{IndexDequeue} does $O(1)$ steps in each level which takes $O(\log p)$ steps. If the response to the \nf{Dequeue} is \nf{null}, \nf{FindResponse} returns \nf{null} in $O(1)$ steps. Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DoublingSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma \ref{dsearchTime}, which is $O(\log q_e+\log q_d)$. Each search in \nf{GetEnqueue()} takes $O(\log p)$ steps since there are at most $p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
\end{proof}

@@

\begin{corollary}
    If the number of concurrent processes in an execution is less than $c$, then each \nf{GetEnqueue} takes $O(\log c)$ steps because each block has at most $c$ operations. So the total amortized worst-case step complexity would be $O(\log p\log c + \log q)$. Furthermore, in a sequential execution where $c=1$, the step complexity of our algorithm would be $\Theta(\log p + \log q)$.
\end{corollary}

>>>>

\vspace{3em}
\textcopyright \; Hossein Naderibeni, 2022

@@

\section*{Acknowledgements}
\hspace{\parindent}First, I want to thank my supervisor Eric Ruppert. Without his guidance and support, this thesis probably would not have been completed. He helped me in every step of this thesis, taught me the shared data structures, helped me find the problem, shape up the algorithm, and write it with its complications because of lots of new notions in it. Thank you again for always pushing to make it better.

I would also like to thank my co-supervisor Franck van Breugel, not only for his careful reading of my thesis and his thoughtful comments but also for his academic advice and motivations.

@@

Last but not the least, I would like to thank my parents for providing me with unfailing support and continuous encouragement throughout my years of study. I want to thank my brother for being there whenever I needed help.

@@

I dedicate this paper to all the people who are fighting for their freedom.

@@

\pagenumbering{arabic}
\section{Introduction}
Shared data structures have become an essential field in distributed algorithms research.
We are reaching the physical limits of how many transistors we can place on a CPU core. The industry solution to provide more computational power is to increase the number of cores of the CPU. It is not hard to see why multiple processes cannot share a sequential data structure. For example, consider two processes trying to append to a sequential linked list simultaneously. Processes $P,Q$ read the same tail node, $P$ changes the next pointer of the tail node to its new node, and then $Q$ does the same. In this run, $P$'s update is overwritten, which makes queue inconsistent. One solution is to use locks; whenever a process wants to update or query a data structure, the process locks the data structure, and others have to wait until the lock is released to read or update the data structure. Using locks has some disadvantages; for example, one process might be slow, and holding a lock for a long time prevents other processes from progressing. Moreover, locks do not allow complete parallelism since only the process holding the lock can progress when the data structure is locked. For these reasons, we are interested in the design of an efficient shared queue without using locks.

@@

Nikolaev and Ravindran~\cite{10.1145/3490148.3538572} present a wait-free queue that uses the fast-path slow-path methodology introduced by Kogan and Petrank~\cite{10.1145/2370036.2145835}. Their work is based on a circular queue using bounded memory. When a process wishes to do an enqueue or a dequeue, it starts two paths. The fast path  ensures good performance while the slow path ensures termination. They show that these two paths do not affect each other and the queue remains consistent. If a process makes no progress, other processes help its slow path to finish. The helping phase suffers from the CAS Retry Problem, because processes compete in a \nf{CAS} loop to decide which succeeds to help. Because of this, the worst-case complexity cannot be better than $\Omega(p)$.

@@

\subsection{Pseudocode} \label{algQ}
We present our algorithm in pseudocode. Page~22 contains the description of the fields in the tree nodes and the blocks. Value of any uninitialized field is \nf{null}. Page~23 contains major routines and the rest of this section consists of the auxiliary routines
The abbreviations bellow are used in the pseudocode and the proof of correctness.
\begin{itemize}
 \item \tt{blocks[$b$].sum\sub{x}=blocks[$b$].sum\sub{x-left}+blocks[$b$].sum\sub{x-right}} (for internal blocks where $b\geq 0$ and \tt{x} $\in$ \{enq, deq\})
%  \item \tt{blocks[b].sum=blocks[b].sum\sub{enq}+blocks[b].sum\sub{deq}}  \tt{ (for b$\geq$0})
  \item \tt{blocks[$b$].num\sub{x}=blocks[$b$].sum\sub{x}-blocks[$b-1$].sum\sub{x}}  (for all blocks where $b>0$ and \tt{x} $\in$ \tt{\{enq, deq, enq-left, enq-right, deq-left, deq-right\}})
\end{itemize}

@@

Now we define the ordering of operations stored in each node. In the non-root nodes, we only need to order operations of a type among themselves that is, we order the \nf{Enqueue}s in the node and order the \nf{Dequeue}s in the node separately. Processes are numbered from 1 to $p$, and leaves of the tree are assigned from left to right. We will show in Lemma \ref{oneOpinBlock} that there is at most one operation from each process in a given block.
\begin{definition} [Ordering of operations inside the nodes] \label{ordering}
\end{definition}
\begin{itemize}
  \item $E(n,b)$ is the sequence of enqueue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows. $E(leaf,b)$ is the single enqueue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents a dequeue operation. If $n$ is an internal node, then
\begin{align*} 
E(n,b) =&  E(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots E(n\nf{.left},n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
&E(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots E(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
\end{align*}
  \item $E_i(n,b)$ is the $i$th enqueue in $E(n,b)$.
\item The order of the enqueue operations in the node $n$ is $E(n)=E(n,1)\cdot E(n,2)\cdot E(n,3)\cdots$
\item $E_i(n)$ is the $i$th enqueue in $E(n)$.
  \item $D(n,b)$ is the sequence of dequeue operations in $ops(n\nf{.blocks[}b\nf{]})$ defined recursively as follows. $D(leaf,b)$ is the single dequeue operation in $ops(leaf\nf{.blocks[}b\nf{]})$ or an empty sequence if $leaf\nf{.blocks[}b\nf{]}$ represents an enqueue operation. If $n$ is an internal node, then
\begin{align*} 
D(n,b) =&  D(n\nf{.left},n\nf{.blocks[}b-1\nf{].end\sub{left}}+1)\cdots D(n\nf{.left}, n\nf{.blocks[}b\nf{].end\sub{left}})\cdot \\ 
&D(n\nf{.right},n\nf{.blocks[}b-1\nf{].end\sub{right}}+1)\cdots D(n\nf{.right},n\nf{.blocks[}b\nf{].end\sub{right}}).
\end{align*}
    \item $D_i(n,b)$ is the $i$th enqueue in $D(n,b)$.
\item The order of the dequeue operations in the node $n$ is $D(n)=D(n,1)\cdot D(n,2)\cdot D(n,3)...$
\item $D_i(n)$ is the $i$th dequeue in $D(n)$.
\end{itemize}

@@

\begin{lemma} \label{doubleRefresh}
  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh} by a process on internal node $n$. If neither $R_1$ nor $R_2$ is a successful \nf{Refresh}, then we have $EST_{n\nf{.left}}^{t^{R{_1}}} \; \cup \; EST_{n\nf{.right}}^{t^{R{_1}}} \subseteq EST_{n}^{^{R{_2}}t}$.
\end{lemma}
\begin{proof}
Let $R_1$ read $i$ from $n\nf{.head}$ at Line \ref{readHead}. By Lemma \ref{lem::headInc}, $R_1$ and $R_2$ cannot both read the same value $i$. By Observation \ref{nonDecreasingHead}, $R_2$ reads a larger value of $n$\nf{.head} than $R_1$.

@@

\begin{lemma}\label{oneOpinBlock}
Each block contains at most one operation of each process.
\end{lemma}
\begin{proof}
To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $P$ in block $B$ in node $n$. Without loss of generality $op_1$ is invoked earlier than $op_2$. Process $P$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins. By Corollary \ref{lem::appendExactlyOnce}, before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $P$'s leaf to the root. Since $b$ contains $op_2$, it must be created after $op_2$ is invoked. The fact that $op_2$\nf{.Append} is invoked after $op_1$\nf{.Append} terminated means that there is some block $B^\prime$ in $n$ before $B$ that contains $op_1$. The existence of $op_1$ in $B$ and $B^\prime$ contradicts Lemma~\ref{lem::noDuplicates}.
\end{proof}

\begin{corollary}\label{blockSize}
    Each block contains at most $c$ operations in total, where $c$ is the number of concurrent processes ($c\leq p$).
\end{corollary}

\begin{lemma}\label{subBlocksBound}
Each block has at most $c$ direct subblocks ($c\leq p$).
\end{lemma}
\begin{proof}
From Definition \ref{def::ops} we know the operations in a block are the union of the operations in the direct subblocks of the block. We can see that each block appended to an internal node contains at least one operation due to the test on Line~\ref{addOP}. Also, blocks in the leaves contain only one \nf{Enqueue} or \nf{Dequeue} operation. By Corollary \ref{blockSize} each block in an internal node contains at most $c$ operations and each one of its direct subblocks has at least one operation, so by pigeonhole principle the number of direct subblocks in a block is at most $c$.
\end{proof}

@@

\begin{lemma}[\nf{DoublingSearch} correctness]\label{dsearch}
If $1\leq e \leq \nf{root.blocks[$end$].sum\sub{enq}}$, then \nf{Doubling-}\\\nf{Search($e$, $end$)} returns \nf{<$b$, $i$>} such that $E_i(root,b)= E_e(root)$.
\end{lemma}
\begin{proof}
 From Lines \ref{setSumEnqLeft} and \ref{setSumEnqRight} we know the \nf{sum\sub{enq-left}}, and \nf{sum\sub{enq-right}} fields of \nf{blocks} in each node are sorted in non-decreasing order. Since $\nf{sum\sub{enq}}=\nf{sum\sub{enq-left}}+ \nf{sum\sub{enq-right}}$, the $\nf{sum\sub{enq}}$ values of \nf{root.blocks[$0\cdot\cdot end$]} are also non-decreasing. By Corollary~\ref{lem::sumX} we know that the \nf{sum\sub{enq}} field in a block is the sum of the number of \nf{Enqueue} operations in that block and the prefix blocks of that block in the node. Furthermore, since $\nf{root.blocks[$0$].sum\sub{enq}}=0$ and $\nf{root.blocks[$end$].sum\sub{enq}} \geq e$, there is a $b$ such that $\nf{root.blocks[$b-1$].sum\sub{enq}}<e$ and $e\leq\nf{root.blocks[$b$].sum\sub{enq}}$. Block \nf{root.blocks[$b$]} contains $E_i(root,b)$.  Lines \ref{dsearchStart}--\ref{dsearchEnd} doubles the search range in Line \ref{doubling} and will eventually reach \nf{start} such that $\nf{root.blocks[start].sum\sub{enq}}\leq e \leq\nf{root.blocks[end].sum\sub{enq}}$. Then, in Line \ref{dsearchBinarySearch}, the binary search finds the $b$ such that \nf{root.blo- cks[$b-1$].sum\sub{enq}$<e\leq$root.blocks[$b$].sum\sub{enq}}. By Corollary \ref{lem::sumX}, \nf{root.blocks[$b$]} is the block that contains $E_e(root)$. Finally, $i$ is computed using the definition of \nf{sum\sub{enq}} and Corollary \ref{lem::sumX}.
\end{proof}

@@

  \caption[Number of \nf{Dequeue} operations before a \nf{Dequeue} in a right child]{The number of \nf{Dequeue} operations before $D_i(n,b)$ shown in the case where $n$ is a right child. The index of the superblock is shown with $sb$. }\label{fig::blockSumDeqRight}
\end{figure}

@@

In the next lemma, we show that the \nf{size} field in each \nf{root block} is computed correctly.
\begin{lemma}\label{sizeCorrectness}
  \nf{root.blocks[$b$].\size}is the \size of the queue after the operations in \nf{root.blocks[$0 \cdot$\\$\cdot\cdot b$]} are applied in the order of~$L$.  
\end{lemma}
\begin{proof}
We prove the claim by induction on $b$. The base case when ${b=0}$ is trivial since the queue is initially empty and $\nf{root.blocks[}0\nf{]}$ contains an empty block with \nf{size} field equal to $0$. We are going to show the correctness when $b=i$ assuming  correctness when $b=i-1$. By Definition \ref{ordering} \nf{Enqueue} operations come before \nf{Dequeue} operations in a block in $L$. By Lemma~\ref{lem::numX} \nf{num\sub{enq}} and \nf{num\sub{deq}} fields in a block show the number of \nf{Enqueue} and \nf{Dequeue} operations in it. If there are more than $\nf{root.blocks[}i-1\nf{].size}+\nf{root.blocks[}i\nf{].num\sub{enq}}$ dequeue operations in \nf{root.blocks[$i$]} then the queue would become empty after \nf{root.blocks[$i$]}. Otherwise, the size of the queue after the $b$th block in the root is $\nf{root.blocks[}b-1\nf{].size}+ \nf{root.blocks[}b\nf{].num\sub{enq}}- \nf{root.blocks[}b\nf{].num\sub{deq}}$. In both cases, this is the same as the assignment on Line \ref{computeLength}.
\end{proof}

@@

\begin{lemma}\label{nullReturn}
$Resp(D_i(\nf{root},b))$ is \nf{null} iff \nf{root.blocks[$b-1$].\size $+$ root.blocks[$b$].num\sub{enq}$- i$ $<0$}.
\end{lemma}
\begin{proof}
The claim follows immediately from Corollary  \ref{numNullDeqBlock} and Lemma \ref{lem::numX}.
\end{proof}

@@

A \nf{Dequeue} creates a block whose \nf{element} is \nf{null}, appends it to the tree, computes its rank among non-null dequeues, finds the corresponding enqueue and returns the response. The first two parts are similar to an \nf{Enqueue} operation and take $O(\log p)$ steps. To compute the rank of a \nf{Dequeue} in $D(n)$, the \nf{Dequeue} calls \nf{IndexDequeue()}. \nf{IndexDequeue} does $O(1)$ steps in each level which takes $O(\log p)$ steps. If the response to the \nf{Dequeue} is \nf{null}, \nf{FindResponse} returns \nf{null} in $O(1)$ steps. Otherwise, if the response to a dequeue in \nf{root.blocks[end]} is in \nf{root.blocks[b]} the \nf{DoublingSearch} takes $\Theta(\log$(\nf{root.blocks[b].size+root.blocks} \nf{[end].size}) by Lemma~\ref{dsearchTime}, which is $O(\log q_e+\log q_d)$. Each search in \nf{GetEnqueue()} takes $O(\log p)$ steps since there are at most $p$ subblocks in a block (Lemma \ref{subBlocksBound}), so \nf{GetEnqueue()} takes $O(\log^2 p)$ steps.
\end{proof}

@@

\begin{observation}
    If the number of concurrent processes in an execution is $c$, then the total amortized worst-case step complexity would be $O(\log p\log c + \log q)$. Furthermore, in a sequential execution where $c=1$, the step complexity of our algorithm would be $\Theta(\log p + \log q)$.
\end{observation}
\begin{proof}
    The analysis is similar ro two previous Lemmas, but by Lemma \ref{subBlocksBound} each \nf{GetEnqueue} takes $O(\log c)$ steps.
\end{proof}
